The 2018 heatwave and its implications on ozone induced damage on vegetation in a subarctic climate

### 2020-04-07 VegOzone.tex
Finished most of Sec 3, only ozone anomalies missing; may need to do the same plot as for the other variables and place the ditailed histograms in an appendix or supplement. Spell-check done.

\introduction  %% \introduction[modified heading if necessary]
\label{sec:intro}
\begin{itemize}
\item ozone in general
\item ozone induced damage in vegetation \emph{Contributor: Ane - maybe also ask Aud?}
  Ozone acts as an oxidative stress to plants. Its main action is imposed through reactions occurring in the cell walls and cell membranes of mesophyll cells inside the leaves.Ozone enters the leaves through stomata, that are kept open to allow for gas exchange between the leaf air spaces and the exterior air. When the stomata are closed, ozone does not enter.
\item metrics of ozone: aot40 and pod
\item present subarctic climate and conditions
  \begin{itemize}
  \item[$\rightarrow$] midnight sun and 23\,\unit{h} open stomata (biological details will be published in a component paper)
  \item[$\rightarrow$] ozone climatology and spring peak
  \end{itemize}
\item future climate and conditions
  \begin{itemize}
  \item[$\rightarrow$] increase in heatwaves and dry periods
  \item[$\rightarrow$] shift in start of growing season (into vicinity of ozone spring peak)
  \item[$\rightarrow$] increase in tropospheric background ozone
  \end{itemize}
\item meteorological situation 2018 vs 2019
\end{itemize}

@@

In our assessment of ozone damage on vegetation in the subarctic in relation to the 2018 heatwave, we focus on the NIBIO operated local site Svanhovd which lies close by the settlement of Svanvik in the Pasvik valley (Section~\ref{subsec:svanhovd}). Throughout the paper, we will use Svanvik synonymously for the atmospheric monitoring site at Svanhovd. The available meteorological monitoring data are described in Section~\ref{subsubsec:atmo_svanvik}. In Section~\ref{subsubsec:ozone_garden}, the ozone garden -- a plant-based ozone monitoring -- will be described in more detail. Other data used for evaluation of the ozone data taken at Svanhovd is described in Section~\ref{subsec:other_data}. We compare with observational data from other monitoring sites in northern Fennoscandia (Section~\ref{subsubsec:ebas}) and use as well reanalysis products (Section~\ref{subsubsec:ozone_rea}) as complement when observational data was not available.

\subsection{NIBIO research station Svanhovd}
\label{subsec:svanhovd}
"Svanhovd is located beside the Pasvik river, in the middle of the wedge of Norwegian land separating Russia and Finland in the north. It is where the eastern Siberian taiga meets the western Boreal forest. [...] Svanhovd was originally established as a State Demonstration and Experimental Farm in the 1930's, in order to expand agriculture in Pasvik valley [...]" - \href{https://www.nibio.no/en/about-eng/addresses/northern-norway/svanhovd}{NIBIO web page, accessed March 2020}.\\
The research station at Svanhovd comprises long-term observations of various agrameteorological variables, i.a. $2\,\unit{m}$ temperature, precipitation, global irradiance, and soil temperature at different depths.

For the purpose of relating observed damage on ozone sensitive plants (e.g. clover and tobacco) an ozone monitor had been installed during the growing seasons 2018/19. The relevant data are described in more detail in Section~\ref{subsubsec:atmo_svanvik}.

The locations of the atmospheric monitoring site as well as the ozone garden with respect to the main buildings of the NIBIO research station are marked in the aerial photography shown in Fig.~\ref{fig:svanhovd_research_station}. 
\emph{Comment: Invite co-authors from NILU and NIBIO?}
     {\bf TODO: Some more general information about the research station? E.g., what NIBIO us currently researching there, that it also hosts a conference center and national park center.}
     "The close proximity to Russia was a dominant factor when the new centre was built in 1992 and established as Svanhovd Environment Centre. The centre building has later been extended and modernised. It contains offices, a conference centre, a national park centre, a brown bear exhibition and a modern laboratory which is also accredited for DNA analysis of brown bear. The centre is surrounded by a botanical garden, one of three such gardens open to the public north of the Polar Circle.

     NIBIO Svanhovd in Pasvik valley is strongly committed to environmental research cooperation in the Barents region. The station disseminates knowledge and performs R\&D based upon the natural environment of this region. It also runs an up-to date DNA-laboratory, a national park centre, a botanical garden and a conference centre. Svanhovd has extensive cooperation with all countries in the Barents region."  - \href{https://www.nibio.no/en/about-eng/addresses/northern-norway/svanhovd}{NIBIO web page, accessed April 2020}.\\

\begin{figure}[t]
  \includegraphics[width=8.3cm]{svanhovd_researchstation}
  \caption{NIBIO research station Svanhovd close by the settlement of Svanvik, Norway. Atmospheric monitoring site and ozone garden have been marked. Aerial photography \copyright Norges Kartverk.}
  \label{fig:svanhovd_research_station}
\end{figure}

@@

In 2018, global irradiance $Q_0$ (Fig.~\ref{fig:data_svanvik_2018_2019}d) displays more sequentially high values in spring (May) and summer (July) compared to 2019, while June 2019 saw more consecutive high irradiance than 2018.

@@

\subsubsection{Surface ozone observations}
\label{subsubsec:ebas}
All surface ozone observations used in this study are available from the EBAS database operated by NILU (\href{http://ebas.nilu.no/}{ebas.nilu.no}). In Table~\ref{tab:ebas_obs}, we give an overview over the exact locations and time range of operational data taking. All stations with available long term observation are locate on higher ground than Svanhovd which will result in higher surface ozone concentrations compared to data taken at Svanvik as ozone abundance decreases with decreasing altitude in the absence of local precursors. {\bf TODO: Citation!} {\bf TODO: The following has to be confirmed and elaborated on!} The vegetation at all sites is similar and consists mainly of pine forests, birch, and heath.

Data from before the 1990s is available solely from Svanvik and Karasjok. These data, however, do not follow the high quality standard procedures implemented nowadays and have to be treated with care {\bf TODO: citation NILU ozone report 200x}. Especially, ozone monitors did not undergo a regular re-calibration. This leads likely to drifts in the observed data and may impose a false periodicity. In 1997, the ozone monitoring at Jergul was relocated downstream the river Karasjokka to a site close by Karasjok. Although the same equipment has been used, this might have introduced a shift in the data series denotes as Jergul/Karasjok in the following. Only few data have been taken upstream the Pasvik river at Janiskovski, hence these will not be taken into further consideration. As can be seen in Fig.~\ref{fig:ozone_timesseries_fenoscandic_obs}, there is an overlap of about $4\,\unit{years}$ between data taken at Esrange, Jergul/Karasjok, and Svanvik. The hatched areas herein indicate the time periods with insufficient quality control as mentioned above.

@@

A climatology derived for global irradiance $Q_0$ is shown in Fig.~\ref{fig:global_rad_clim}a) for both, climatological average of hourly maximum $\left<Q_0^\mathrm{max}\right>$, minimum $\left<Q_0^\mathrm{min}\right>$, and mean $\left<Q_0^\mathrm{mean}\right>$ irradiance observed at Svanvik. In accordance to the geometrically derived sunshine duration shown in Fig.~\ref{fig:sunlight_fennoscandia}, $\left<Q_0^\mathrm{max}\right>$ approaches zero in December/January commonly known as polar night and reaches its maximum in June/July commonly referred to as midnight sun conditions. The highest observed $\left<Q_0^\mathrm{max}\right>$ is about $800\,\unit{W\,m^{-2}}$, while on average $\left<Q_0^\mathrm{mean}\right>$ it is about $450\,\unit{W\,m^{-2}}$ (regularly reached in May--July). The maximum of the climatological minimum $\left<Q_0^\mathrm{min}\right>$ lies at about $200\,\unit{W\,m^{-2}}$.

@@

The correlation of \chem{[O_3]} between Esrange, Pallas, and Jergul/Karasjok is fairly high ($r^2\approx 0.8$, Fig.~\ref{fig:density_distribution}a, b), hence we can put those data series together to derive a general ozone climatology for northern Fennoscandia. In Figure~\ref{fig:ozone_climatology_fenoscandic_obs_spline}, the 2 dimensional density distributions for the thus defined northern Fennoscandia (Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_spline}a) and Svanvik (Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_spline}b) are shown. Darker colors indicate higher probability to observe these values. On top of the density distributions, $10\,\unit{days}$ averages of daily mean \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}}, daily maximum \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, max}}, and daily minimum \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, min}} are displayed together with the $1 \sigma$ uncertainties, respectively. In addition, \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}} is also shown with standard error of mean (albeit small and hardly visible). Splines have been fitted through the data for easier inter-comparison with reanalysis data.

From the northern Fennoscandic climatology, one can deduce several features. The maximum of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, min}} peaks about $20\,\unit{days}$ earlier than the \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}}, and \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, max}}. With respect of the periodicity of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}}, it takes about $8.5\,\unit{months}$ from minimum concentration to maximum concentration, while from maximum to minimum it only takes about $4.5\,\unit{months}$. The maximum of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, max}} is at $52\,\unit{ppb}$ in late April and the minimum of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, min}} occurs at $15\,\unit{ppb}$ in the beginning of September. The steep rise of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}} in spring coincident with increasing temperatures as well as increase in global irradiance. The decrease in \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}} coincides with the estimated onset of the greening season based on the agricultural rule of thumb that the daily average temperature has to be above $5\,\unit{^\circ C}$ for $5\,\unit{days}$ consecutively (day of the year 110--150, Fig~\ref{fig:greening_season_change_Svanvik}).

As mentioned above, the average ozone concentration at Svanvik is in general lower than at the other sites which is in league with a decrease in tropospheric ozone with decreasing altitude. The density distribution for Svanvik and the respective \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}}, and \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, max}} (Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_spline}b) display a similar pattern as the northern Fennoscandic climatology. Only does the spring peak apparently occur slightly earlier (in the beginning of April).

@@

Finally, we shall look at the comparison of climatologies derived from observational data ozone and the various reanalysis products (Fig.~\ref{fig:ozone_climatologies}). As representative for the sites, the nearest neighboring grid point has been chosen. The ECMWF products do not reproduce observed surface ozone well. The MACC reanalysis is in general too low and displays a erroneous seasonal cycle. The CAMS reanalysis matches better but does not reproduce the seasonality in northern Fennoscandia. The ensemble mean of the Copernicus regional model reanalysis does reproduce the seasonal cycle in northern Fennoscandia, though the modeled ozone concentrations in the north are slightly to low.

\begin{figure*}[t]
  \includegraphics[width=12cm]{ozone_climatologies}
  \caption{Comparison of climatologies derived from observational data (EBAS, see Fig.~\ref{fig:ozone_climatology_fenoscandic_obs}) with various ozone reanalysis products: (a) ECMWF: MACC reanalysis; (b) ECMWF: CAMS reanalysis; (c) Copernicus: Regional model reanalysis ensemble mean. As representative for the sites, the nearest neighboring grid point has been chosen. The ECMWF products do not reproduce observed surface ozone well. The MACC reanalysis is in general too low and displays a erroneous seasonal cycle. The CAMS reanalysis matches better but does not reproduce the seasonality in northern Fennoscandia. The ensemble mean of the Copernicus regional model reanalysis does reproduce the seasonal cycle in northern Fennoscandia, though the modeled ozone concentrations in the north are slightly to low.}
  \label{fig:ozone_climatologies}
\end{figure*}

\subsection{2018/19 anomalies}
\label{subsec:anomalies}

\subsubsection{Temperature, precipitation, and global irradiance}
\label{subsubsec:anomal_tpq}

\begin{figure}[t]
  \includegraphics[width=8.3cm]{temperature_signific}
  \caption{Temperature at Svanvik. Percentage of days warmer/colder than $\pm 1\,\sigma$ from climatological mean displayed for each month. Negative deviations from climatology are shown as negative percentage. The annual positive/negative deviations are indicated in the respective corners (right upper/right lower). The dashed lines indicate the expected percentage of values falling above/below $\pm 1\,\sigma$ if a normal distribution is assumed ($15.9\,\unit{\%}$). Year displayed in panel (a) 2018; (b) 2019. The summer of 2018 was significantly warmer than average, especially May and July. The summer of 2019 was fairly average but significantly warmer than average in April and colder in July/August.}
  \label{fig:plot_temperature_anomalies_svanvik}
\end{figure}

\begin{figure}[t]
  \includegraphics[width=8.3cm]{precipitation_signific}
  \caption{Precipitation at Svanvik. Percentage of days wetter/drier than $\pm \frac{1}{2}\,\sigma$ from climatological mean displayed for each month. Negative deviations from climatology are shown as negative percentage. The annual positive/negative deviations are indicated in the respective corners (right upper/right lower). The dashed lines indicate the expected percentage of values falling above $\pm\frac{1}{2}\,\sigma$ if a normal distribution is assumed ($30.8\,\unit{\%}$). Year displayed in panel (a) 2018; (b) 2019. The summer of 2018 was significantly drier than in 2019.}
  \label{fig:plot_precipitation_anomalies_svanvik}
\end{figure}

\begin{figure}[t]
  \includegraphics[width=8.3cm]{global_rad_signific}
  \caption{Global irradiance at Svanvik. Percentage of days above/below $\pm 1\,\sigma$ from climatological mean displayed for each month. Negative deviations from climatology are shown as negative percentage. The annual positive/negative deviations are indicated in the respective corners (right upper/right lower). The dashed lines indicate the expected percentage of values falling above/below $\pm 1\,\sigma$ if a normal distribution is assumed ($15.9\,\unit{\%}$). Year displayed in panel (a) 2018; (b) 2019. In summer 2018 (May/July), global irradiance was significantly higher than on average, while 2019 was generally an average year.}
  \label{fig:global_rad_signific}
\end{figure}

\subsubsection{Ozone}
\label{subsubsec:anomal_ozone}

>>>>

\introduction  %% \introduction[modified heading if necessary]
\label{sec:intro}
\begin{itemize}
\item ozone in general
\item ozone induced damage in vegetation \emph{Contributor: Ane - maybe also ask Aud?}
  Ozone acts as an oxidative stress to plants. Its main action is imposed through reactions occurring in the cell walls and cell membranes of mesophyll cells inside the leaves.Ozone enters the leaves through stomata, that are kept open to allow for gas exchange between the leaf air spaces and the exterior air. When the stomata are closed, ozone does not enter.
\item metrics of ozone: aot40 and pod
\item present subarctic climate and conditions
  \begin{itemize}
  \item[$\rightarrow$] midnight sun and 23\,\unit{h} open stomata (biological details will be published in a component paper)
  \item[$\rightarrow$] ozone climatology and spring peak {\bf TODO: Cite review article - find newer articles.}
  \end{itemize}
\item future climate and conditions
  \begin{itemize}
  \item[$\rightarrow$] increase in heatwaves and dry periods
  \item[$\rightarrow$] shift in start of growing season (into vicinity of ozone spring peak)
  \item[$\rightarrow$] increase in tropospheric background ozone
  \end{itemize}
\item meteorological situation 2018 vs 2019
\end{itemize}

@@

In our assessment of ozone damage on vegetation in the subarctic in relation to the 2018 heatwave, we focus on the NIBIO operated Environment Centre Svanhovd which lies close by the settlement of Svanvik in the Pasvik valley (Section~\ref{subsec:svanhovd}). Throughout the paper, we will use Svanvik synonymously for the atmospheric monitoring site at Svanhovd. The available meteorological monitoring data are described in Section~\ref{subsubsec:atmo_svanvik}. In Section~\ref{subsubsec:ozone_garden}, the ozone garden -- a plant-based ozone monitoring -- will be described in more detail. Other data used for evaluation of the ozone data taken at Svanhovd is described in Section~\ref{subsec:other_data}. We compare with observational data from other monitoring sites in northern Fennoscandia (Section~\ref{subsubsec:ebas}) and use as well reanalysis products (Section~\ref{subsubsec:ozone_rea}) as complement when observational data was not available.

\subsection{NIBIO Environment Centre Svanhovd}
\label{subsec:svanhovd}
"Svanhovd is located beside the Pasvik river, in the middle of the wedge of Norwegian land separating Russia and Finland in the north. It is where the eastern Siberian taiga meets the western Boreal forest. [...] Svanhovd was originally established as a State Demonstration and Experimental Farm in the 1930's, in order to expand agriculture in Pasvik valley [...]" - \href{https://www.nibio.no/en/about-eng/addresses/northern-norway/svanhovd}{NIBIO web page, accessed March 2020}.

The locations of the atmospheric monitoring site as well as the ozone garden with respect to the main buildings of the Svanhovd Environment Centre are marked in the aerial photography shown in Fig.~\ref{fig:svanhovd_research_station}. The Centre is located south of the settlement of Svanvik, where as the border to Russia, the Pasvik river, lies only a few hundred meters to the east. The original farm buildings, protected by law, are located in the lower center, while the Svanhovd Environment Centre follows to the north-east. The Centre hosts offices, a conference center, a national park center, a brown bear exhibition as well as a modern laboratory for DNA analysis. Left of the Centre and north of the original farm buildings a botanical garden can be found. It is one amongst three such gardens north of the Polar Circle. The Svanhovd Environment Centre focuses on research of the special environment of the region, but also comprises long-term observations of various agrameteorological variables, i.a. $2\,\unit{m}$ temperature, precipitation, global irradiance, and soil temperature at different depths.

For the purpose of relating observed damage on ozone sensitive plants (e.g. clover and tobacco) to local ozone concentrations, an ozone monitor had been installed during the growing seasons 2018/19. The relevant data are described in more detail in Section~\ref{subsubsec:atmo_svanvik}.

\emph{Comment: Invite co-authors from NILU and NIBIO?}

\begin{figure}[t]
  \includegraphics[width=8.3cm]{svanhovd_researchstation}
  \caption{NIBIO Environment Centre Svanhovd close by the settlement of Svanvik, Norway. Atmospheric monitoring site and ozone garden have been marked. Aerial photography \copyright Norges Kartverk.}
  \label{fig:svanhovd_research_station}
\end{figure}

@@

In 2018, global irradiance $Q_0$ (Fig.~\ref{fig:data_svanvik_2018_2019}d) displays more sequentially high values in spring (May) and summer (July) compared to 2019, while June 2019 saw more consecutive high irradiance than 2018. In both years, the maximum recorded $Q_0$ was roughly $750\,\unit{W\,m^{-2}}$ reached only in June.

@@

\subsubsection{Surface ozone observations}
\label{subsubsec:ebas}
All surface ozone observations used in this study are available from the EBAS database operated by NILU (\href{http://ebas.nilu.no/}{ebas.nilu.no}). In Table~\ref{tab:ebas_obs}, we give an overview over the exact locations and time range of operational data taking. All stations with available long term observations are locate on higher ground than Svanhovd which will result in higher surface ozone concentrations in comparison as ozone abundance, in general, decreases with decreasing altitude in the absence of local precursors. {\bf TODO: Citation!} {\bf TODO: The following has to be confirmed and elaborated on! Ane?} The vegetation at all sites is similar and consists mainly of pine forests, birch, and heath.

Data from before the 1990s is available solely from Svanvik and Karasjok. These data, however, do not follow the high quality standard procedures implemented nowadays and have to be treated with care {\bf TODO: citation NILU ozone report 200x}. Especially, ozone monitors did not undergo a regular re-calibration. This leads likely to drifts in the observed data and may impose a false periodicity. In 1997, the ozone monitoring at Jergul was relocated downstream the river Karasjokka to a site close by Karasjok. Although the same equipment has been used, this might have introduced a breaking point in the combined data series denoted as Jergul/Karasjok in the following which would be of a problem if this was the sole data to compare with. Only few data have been taken upstream the Pasvik river at Janiskovski, hence these will not be taken into further consideration, but are shown for completion. As can be seen in Fig.~\ref{fig:ozone_timesseries_fenoscandic_obs}, there is an overlap of about $4\,\unit{years}$ between data taken at Esrange, Jergul/Karasjok, and Svanvik. The hatched areas indicate time periods with insufficient quality control as mentioned above.

@@

A climatology derived for global irradiance $Q_0$ is shown in Fig.~\ref{fig:global_rad_clim}a) for both, climatological average of hourly maximum $\left<Q_0^\mathrm{max}\right>$, minimum $\left<Q_0^\mathrm{min}\right>$, and mean $\left<Q_0^\mathrm{mean}\right>$ irradiance observed at Svanvik. In agreement with the geometrically derived sunshine duration shown in Fig.~\ref{fig:sunlight_fennoscandia}, $\left<Q_0^\mathrm{max}\right>$ approaches zero in December/January commonly known as polar night and reaches its maximum in June/July commonly referred to as midnight sun conditions. The highest observed $\left<Q_0^\mathrm{max}\right>$ is about $800\,\unit{W\,m^{-2}}$, while on average $\left<Q_0^\mathrm{mean}\right>$ it is about $450\,\unit{W\,m^{-2}}$ (regularly reached in May--July). The maximum of the climatological minimum $\left<Q_0^\mathrm{min}\right>$ lies at about $200\,\unit{W\,m^{-2}}$.

@@

The correlation of \chem{[O_3]} between Esrange, Pallas, and Jergul/Karasjok is fairly high ($r^2\approx 0.8$, Fig.~\ref{fig:density_distribution}a, b), hence we can put those data series together to derive a general ozone climatology for northern Fennoscandia. The correlation between Svanvik and Esrange and Pallas is rather low, 0.4 and 0.6, respectively. Thus, Svanvik cannot directly be represented by any measurements taken at either Esrange or Pallas.

In Figure~\ref{fig:ozone_climatology_fenoscandic_obs_spline}, the 2 dimensional density distributions for the thus defined northern Fennoscandia (Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_spline}a) and Svanvik (Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_spline}b) are shown. Darker colors indicate higher probability to observe these values. On top of the density distributions, $10\,\unit{days}$ averages of daily mean \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}}, daily maximum \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, max}}, and daily minimum \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, min}} are displayed together with the $1 \sigma$ uncertainties, respectively. In addition, \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}} is also shown with standard error of mean (albeit small and hardly visible). Splines have been fitted through the data for easier inter-comparison with reanalysis data.

From the northern Fennoscandic climatology, one can deduce several features. The maximum of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, min}} peaks about $20\,\unit{days}$ earlier than the \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}}, and \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, max}}. With respect to the periodicity of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}}, it takes about $8.5\,\unit{months}$ from minimum concentration to maximum concentration, while from maximum to minimum it only takes about $4.5\,\unit{months}$. The maximum of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, max}} is at $52\,\unit{ppb}$ in late April and the minimum of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, min}} occurs at $15\,\unit{ppb}$ in the beginning of September. The steep rise of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}} in spring coincident with increasing temperatures as well as increase in global irradiance. The decrease in \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}} coincides with the estimated onset of the greening season based on the agricultural rule of thumb that the daily average temperature has to be above $5\,\unit{^\circ C}$ for $5\,\unit{days}$ consecutively (day of the year 110--150 at Svanvik, Fig~\ref{fig:greening_season_change_Svanvik}).

The density distribution for Svanvik and the respective \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}}, and \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, max}} (Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_spline}b) display a similar pattern as the northern Fennoscandic climatology. The spring peak apparently occurs slightly earlier (in the beginning of April). As Svanvik is located lower than the other sites, the earlier ozone peak may point to a strong interplay between ozone uptake by vegetation after the onset of greening, for at higher altitude plants sprout later. The average ozone concentration at Svanvik is lower than at the other sites, as already mentioned, which is in league with a decrease in tropospheric ozone with decreasing altitude. In July--September, ozone is occasionally almost completely depleted during night time.

@@

Finally, we shall look at the comparison of climatologies derived from observational ozone data and the various reanalysis products (Fig.~\ref{fig:ozone_climatologies}). As representative for the sites, the nearest neighboring grid point has been chosen. The ECMWF products do not reproduce observed surface ozone well. The MACC reanalysis (Fig.~\ref{fig:ozone_climatologies}a) is in general too low ($(9\pm 7)\,\unit{ppb}$) and displays a erroneous seasonal cycle. MACC reanalysis ozone concentrations peak already in March followed by a second peak in July. Also CAMSRA matches the observed ozone climatology hardly (Fig.~\ref{fig:ozone_climatologies}b). Despite displaying only one ozone peak, it still does not reproduce the seasonality in northern Fennoscandia at all. The CAMSRA spring peak appears to lag behind by $1\,\unit{month}$ and is $5\,\unit{ppb}$ lower, whereas the minimum occurs in January compared to August/September. The annual amplitude ($(26\pm 1)\,\unit{ppb}$) is larger than in the climatology derived from observations ($19\,\unit{ppb}$). Modeled \chem{[O_3]} at Svanvik are even slightly higher than at Pallas. In contrast, the ensemble mean of the Copernicus regional model reanalysis does reproduce the seasonal cycle in northern Fennoscandia quite well (Fig.~\ref{fig:ozone_climatologies}c), but is on average still slightly too low ($(2.8\pm 0.5)\,\unit{ppb}$).

\begin{figure}[t]
  \includegraphics[width=8.3cm]{ozone_climatologies_portrait}
  \caption{Comparison of climatologies derived from observational data (EBAS, see Fig.~\ref{fig:ozone_climatology_fenoscandic_obs}) with various ozone reanalysis products: (a) ECMWF: MACC reanalysis; (b) ECMWF: CAMS reanalysis; (c) Copernicus: Regional model reanalysis ensemble mean. As representative for the sites, the nearest neighboring grid point has been chosen. The ECMWF products do not reproduce observed surface ozone well. The MACC reanalysis is in general too low and displays a erroneous seasonal cycle. The CAMS reanalysis matches better but does not reproduce the seasonality in northern Fennoscandia. The ensemble mean of the Copernicus regional model reanalysis does reproduce the seasonal cycle in northern Fennoscandia, though the modeled ozone concentrations are slightly too low.}
  \label{fig:ozone_climatologies}
\end{figure}

\subsection{2018/19 anomalies}
\label{subsec:anomalies}
Using the established climatologies, we look at the anomalies in 2018/19 and the differences between the two years.

\subsubsection{Temperature, precipitation, and global irradiance}
\label{subsubsec:anomal_tpq}
In Fig.~\ref{fig:plot_temperature_anomalies_svanvik}, temperature anomalies at Svanvik are shown as percentage of days warmer/colder than $\pm 1\,\sigma$ from the climatological mean for each month. Negative deviations from climatology are displayed as negative percentage. The annual positive/negative deviations are indicated in the respective corners (right upper/right lower). The dashed lines indicate the expected percentage of values falling above/below $\pm 1\,\sigma$ if a normal distribution is assumed ($15.9\,\unit{\%}$). The summer of 2018 was significantly warmer than average. Especially in May and July more than $40\,\unit{\%}$ of days were significantly warmer than normal temperature. Temperatures continued to be significantly warmer throughout fall into early winter (August--November) with the exception of October. March 2018 was unusually cold. The summer of 2019 was fairly average, however significantly colder in July, while August was both significantly colder and warmer. Temperatures in April were significantly warmer than average.

\begin{figure}[t]
  \includegraphics[width=8.3cm]{temperature_signific}
  \caption{Temperature anomalies at Svanvik. Percentage of days warmer/colder than $\pm 1\,\sigma$ from climatological mean displayed for each month. Negative deviations from climatology are shown as negative percentage. The annual positive/negative deviations are indicated in the respective corners (right upper/right lower). The dashed lines indicate the expected percentage of values falling above/below $\pm 1\,\sigma$ if a normal distribution is assumed ($15.9\,\unit{\%}$). Year displayed in panel (a) 2018; (b) 2019. The summer of 2018 was significantly warmer than average, especially May and July. The summer of 2019 was fairly average but significantly warmer than average in April and colder in July/August.}
  \label{fig:plot_temperature_anomalies_svanvik}
\end{figure}

In a similar way, the precipitation anomalies at Svanvik are displayed in Fig.~\ref{fig:plot_precipitation_anomalies_svanvik}. However, the percentage of days wetter/drier than $\pm \frac{1}{2}\,\sigma$ from the climatological mean is shown for each month, due to the larger variability in observed precipitation too few days showed a significance of $1 \sigma$ or higher. As above, negative deviations from the climatology are shown as negative percentage and annual positive/negative deviations are indicated in the respective corners. The dashed lines indicate the expected percentage of values falling above $\pm\frac{1}{2}\,\sigma$ if a normal distribution is assumed ($30.8\,\unit{\%}$). Unlike for the temperature anomalies, the picture is not as clear. While March 2018 was unusually wet, summer and fall (May/July, September/October) were significantly drier than average. 2019 was rather average throughout the year, with drier than usual fall (September/October) and a wet December.

\begin{figure}[t]
  \includegraphics[width=8.3cm]{precipitation_signific}
  \caption{Precipitation anomalies at Svanvik. Percentage of days wetter/drier than $\pm \frac{1}{2}\,\sigma$ from climatological mean displayed for each month. Negative deviations from climatology are shown as negative percentage. The annual positive/negative deviations are indicated in the respective corners (right upper/right lower). The dashed lines indicate the expected percentage of values falling above $\pm\frac{1}{2}\,\sigma$ if a normal distribution is assumed ($30.8\,\unit{\%}$). Year displayed in panel (a) 2018; (b) 2019. The summer of 2018 was significantly drier than in 2019.}
  \label{fig:plot_precipitation_anomalies_svanvik}
\end{figure}

Similarly, the global irradiance anomalies at Svanvik are presented in Fig.~\ref{fig:global_rad_signific}. The percentage of days above/below $\pm 1\,\sigma$ from the climatological mean is displayed for each month. The dashed lines indicate the expected percentage of values falling above/below $\pm 1\,\sigma$ if a normal distribution is assumed. In summer 2018 (May/July), global irradiance was significantly higher than on average, while 2019 was generally an average year with a lightly darker spring (April/May).

\begin{figure}[t]
  \includegraphics[width=8.3cm]{global_rad_signific}
  \caption{Global irradiance anomalies at Svanvik. Percentage of days above/below $\pm 1\,\sigma$ from climatological mean displayed for each month. Negative deviations from climatology are shown as negative percentage. The annual positive/negative deviations are indicated in the respective corners (right upper/right lower). The dashed lines indicate the expected percentage of values falling above/below $\pm 1\,\sigma$ if a normal distribution is assumed ($15.9\,\unit{\%}$). Year displayed in panel (a) 2018; (b) 2019. In summer 2018 (May/July), global irradiance was significantly higher than on average, while 2019 was generally an average year.}
  \label{fig:global_rad_signific}
\end{figure}

In summary, 2018 was significantly warmer, drier, and lighter than the climatological average, while 2019 was a rather average year.

\subsubsection{Ozone}
\label{subsubsec:anomal_ozone}
As we have found above, meteorological conditions in 2018 were significantly above average. How does this translate into observed ozone anomalies?
